---
title: "How do we want to display CATEGORY data?"
author: "Jessica Couture"
date: "5/14/2016"
output: html_document
---

```{r setup, echo=FALSE, warning=FALSE,message=FALSE}
library(knitr)
library(dplyr)
library(readr)
library(ggplot2)
library(reshape2)
library(grid)
library(gridExtra)

rslt=read.csv('data/archResRevised2.csv',header=T,stringsAsFactors=F)
rslt[rslt$Status=='published','Status']<-'Published'

rslt$succ=ifelse(rslt$Status %in% c('Published','SentData','Revised'),'aff','neg')
rslt$plotSucc=ifelse(rslt$Status %in% c('Published','SentData','Revised'),1,2)# 1 = success, 2 = not successful

rslt$statSucc=ifelse(rslt$Status %in% c('Published','SentData','Revised'),1,0)# 1 = success, 0 = not successful


agDes=read.csv('data/ecoInfoAgencyDesig.csv',header=T,stringsAsFactors=T,col.names=c('agency','agGrp','agSubGrp'))
agDes$agGrp=gsub('nonProf','private',agDes$agGrp)

rslt2=merge(rslt,agDes,all.x=T)
rslt2=rslt2 %>%
  filter(!end>2010) 
```

## Emphasize percents? number of datasets? differences between categories?...
We have talked about a number of different ways to display these results. Here are the statistics to show what results we have, followed by the potential visulaizations. We'll use the same format for the sector plots (question 3) so I haven't included those here. 

> Q3. Are there certain **research fields** that are more likely to make data available than others?

```{r stats-fine, echo=F}
dtProps=table(rslt2$succ,rslt2$dataType)
dtChi=chisq.test(dtProps)

dtChi #diff between data fields

succBlr<-glm(formula = statSucc~dataType,family='binomial',data=rslt2)

## statSucc = Binomial: did we recover the data?
## agGrp = sector of home institution
## end = last year project was funded by EVOSTC
## dataType = field of study

summary(succBlr)
```

## Statistical outputs: Broad field of study categories
```{r stats-broad, echo=F}
fieldSpl=strsplit(rslt2$dataType,spl="-")
rslt2$broadDT=sapply(fieldSpl,function(x) x[1])

bdtProps=table(rslt2$succ,rslt2$broadDT)
bdtChi=chisq.test(bdtProps)

bdtChi

succBlr<-glm(formula = statSucc~broadDT,family='binomial',data=rslt2)

summary(succBlr)
```
#### Physical data is the leading indicator of success...

***

## Percentage plots:

```{r nDatasets - broad, echo=FALSE}
yrsCols=c("neg"='grey',
          "aff"='black')

catYN=as.data.frame(table(rslt2$dataType,rslt2$succ))
colnames(catYN)=c('category','succ','nDatasets')

broadSpl=strsplit(as.character(catYN$category),'-')
catYN$broad=sapply(broadSpl,function(x) x[1])
bioSub=sapply(broadSpl,function(x) x[2])

bioYN=catYN %>%
  filter(broad=='biological') %>%
  mutate(subCat=na.omit(bioSub)) %>%
  group_by(category) %>%
  mutate(grpTots=sum(nDatasets)) %>%
  mutate(prop=nDatasets/grpTots)

bioYN$subCat=factor(bioYN$subCat,levels=c('plankton','benthicInverts','fish','birds','mammals'))

##### general fields (biology grouped)

sums=bioYN %>%
  group_by(succ) %>%
  summarise(nDatasets=sum(nDatasets)) %>%
  mutate(category='biological') %>%
  select(category,succ,nDatasets)

broadDf=catYN %>%
  filter(!broad=='biological') %>%
  group_by(broad) %>%
  mutate(grpTots=sum(nDatasets)) %>%
  mutate(prop=nDatasets/grpTots)
  
broadDf$category<-factor(broadDf$category,levels=c('physical','chemical','biological','ecosystem/habitat','social'))
broadDf[9:10,]<-sums
broadDf[9:10,'prop'] <- c(broadDf$nDatasets[9]/sum(sums$nDatasets),broadDf$nDatasets[10]/sum(sums$nDatasets))
broadDf[9:10,'grpTots'] <- sum(sums$nDatasets)

broadSucc=ggplot(broadDf[broadDf$succ==0,],aes(x=category,y=grpTots))+
  geom_bar(stat = "identity",fill="grey")+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  theme(axis.text.x = element_text(angle=90))+
  #ylim=c(0,300)+
  xlab('Field')+
  ylab('Number of datasets\n')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))#+
  # annotate("text", x = 1, y=broadDf$grpTots[[3]]+9, label = as.character(round(broadDf[broadDf$category=="physical" & broadDf$succ=="aff","prop"],2)))+
  # annotate("text", x = 2, y=broadDf$grpTots[[1]]+9, label = as.character(round(broadDf[broadDf$category=="chemical" & broadDf$succ=="aff","prop"],2)))+
  # annotate("text", x = 3, y=broadDf$grpTots[[9]]+9, label = as.character(round(broadDf[broadDf$category=="biological" & broadDf$succ=="aff","prop"],2)))+
  # annotate("text", x = 4, y=broadDf$grpTots[[2]]+9, label = as.character(round(broadDf[broadDf$category=="ecosystem/habitat" & broadDf$succ=="aff","prop"],2)))+
  # annotate("text", x = 5, y=broadDf$grpTots[[4]]+9, label = as.character(round(broadDf[broadDf$category=="social" & broadDf$succ=="aff","prop"],2)))
broadSucc
```

***

``` {r percent successful stacked - broad, echo=FALSE}

broadPerc=ggplot(broadDf[broadDf$succ==0,],aes(x=category,y=(prop*100)))+
  geom_bar(stat = "identity")+
  #scale_fill_manual(values=yrsCols,name='Success?',breaks=c('neg','aff'),labels=c('no','yes'))+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  theme(axis.text.x = element_text(angle=90))+
  ylim(0,75)+
  xlab('Field')+
  ylab('Percent successful')+
  #ggtitle('Percent success by reaserch field')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))+
  annotate("text", x = 1, y=broadDf$prop[[3]]+0.05, label = as.character(broadDf[broadDf$category=="physical" & broadDf$succ==0,"grpTots"]))+
  annotate("text", x = 2, y=broadDf$prop[[1]]+0.05, label = as.character(broadDf[broadDf$category=="chemical" & broadDf$succ==0,"grpTots"]))+
  annotate("text", x = 3, y=broadDf$prop[[9]]+0.05, label = as.character(broadDf[broadDf$category=="biological" & broadDf$succ==0,"grpTots"]))+
  annotate("text", x = 4, y=broadDf$prop[[2]]+0.05, label = as.character(broadDf[broadDf$category=="ecosystem/habitat" & broadDf$succ==0,"grpTots"]))+
  annotate("text", x = 5, y=broadDf$prop[[4]]+0.05, label = as.character(broadDf[broadDf$category=="social" & broadDf$succ==0,"grpTots"]))
broadPerc
```

***

```{r nDatasets original - biology, echo=FALSE}
bioSucc=ggplot(bioYN,aes(x=subCat,y=nDatasets,fill=succ,order=rev(succ)))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values=yrsCols,name='Success?',breaks=c('neg','aff'),labels=c('no','yes'))+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  theme(axis.text.x = element_text(angle=90))+
  #ylim=c(0,300)+
  xlab('Field')+
  ylab('Number of datasets\n')+
  ggtitle('Successes of biology fields')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))+
  annotate("text", x = 1, y=bioYN$grpTots[[5]]+5, label = as.character(round(bioYN[bioYN$category=="biological-plankton" & bioYN$succ=="aff","prop"],2)))+
  annotate("text", x = 2, y=bioYN$grpTots[[1]]+5, label = as.character(round(bioYN[bioYN$category=="biological-benthicInverts" & bioYN$succ=="aff","prop"],2)))+
  annotate("text", x = 3, y=bioYN$grpTots[[3]]+5, label = as.character(round(bioYN[bioYN$category=="biological-fish" & bioYN$succ=="aff","prop"],2)))+
  annotate("text", x = 4, y=bioYN$grpTots[[2]]+5, label = as.character(round(bioYN[bioYN$category=="biological-birds" & bioYN$succ=="aff","prop"],2)))+
  annotate("text", x = 5, y=bioYN$grpTots[[4]]+5, label = as.character(round(bioYN[bioYN$category=="biological-mammals" & bioYN$succ=="aff","prop"],2)))
broadSucc
bioSucc
```

***

```{r percent successful - biologyOrig, echo=FALSE}
bioPerc=ggplot(bioYN[bioYN$succ=='aff',],aes(x=subCat,y=prop,order=rev(succ)))+
  geom_bar(stat = "identity")+
  #scale_fill_manual(values=yrsCols,name='Success?',breaks=c('neg','aff'),labels=c('no','yes'))+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  theme(axis.text.x = element_text(angle=90))+
  #ylim=c(0,300)+
  xlab('Field')+
  ylab('Percent successful')+
  ggtitle('Percent success within the biology field')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))+
  annotate("text", x = 1, y=bioYN$prop[[5]]+0.05, label = as.character(bioYN[bioYN$category=="biological-plankton" & bioYN$succ=="aff","grpTots"]))+
  annotate("text", x = 2, y=bioYN$prop[[1]]+0.05, label = as.character(bioYN[bioYN$category=="biological-benthicInverts" & bioYN$succ=="aff","grpTots"]))+
  annotate("text", x = 3, y=bioYN$prop[[3]]+0.05, label = as.character(bioYN[bioYN$category=="biological-fish" & bioYN$succ=="aff","grpTots"]))+
  annotate("text", x = 4, y=bioYN$prop[[2]]+0.05, label = as.character(bioYN[bioYN$category=="biological-birds" & bioYN$succ=="aff","grpTots"]))+
  annotate("text", x = 5, y=bioYN$prop[[4]]+0.05, label = as.character(bioYN[bioYN$category=="biological-mammals" & bioYN$succ=="aff","grpTots"]))
broadPerc
bioPerc
```

***

```{r percent successful - biologyRevised, echo=FALSE}
bioPerc=ggplot(bioYN[bioYN$succ==0,],aes(x=subCat,y=(prop*100)))+
  geom_bar(stat = "identity",fill="black")+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  theme(axis.text.x = element_text(angle=90))+
  ylim(0,75)+
  xlab('Bio Field')+
  ylab('% successful')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))+
  annotate("text", x = 1, y=bioYN$prop[[5]]*100+5, label = as.character(bioYN[bioYN$category=="biological-plankton" & bioYN$succ==0,"grpTots"]))+
  annotate("text", x = 2, y=bioYN$prop[[1]]*100+5, label = as.character(bioYN[bioYN$category=="biological-benthicInverts" & bioYN$succ==0,"grpTots"]))+
  annotate("text", x = 3, y=bioYN$prop[[3]]*100+5, label = as.character(bioYN[bioYN$category=="biological-fish" & bioYN$succ==0,"grpTots"]))+
  annotate("text", x = 4, y=bioYN$prop[[2]]*100+5, label = as.character(bioYN[bioYN$category=="biological-birds" & bioYN$succ==0,"grpTots"]))+
  annotate("text", x = 5, y=bioYN$prop[[4]]*100+5, label = as.character(bioYN[bioYN$category=="biological-mammals" & bioYN$succ==0,"grpTots"]))
broadPerc
bioPerc
```


***

```{r nDatasets - biologyRevised, echo=FALSE}
bioSucc=ggplot(bioYN[bioYN$succ==0,],aes(x=subCat,y=grpTots))+
  geom_bar(stat = "identity",fill="grey")+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  theme(axis.text.x = element_text(angle=90),
        axis.title.x = element_blank())+
  #ylim=c(0,300)+
  ggtitle('Biology sub-fields')+
  ylab('Number of datasets\n')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))
broadSucc
bioSucc
```

Bubble plots?:
``` {r bubble plots, echo=F}

```