---
title: "EVOSTC data collection paper - Analysis"
author: "Jessica Couture"
date: "9/27/2015"
output: html_document
---

#Background

We are writing up a publication to report results of a two year data archiving effort by a small group of researchers and students at the National Center for Ecological Analysis and Synthesis at UC Santa Barbara. The [Exxon Valdez Oil Spill Trustee Council](http://www.evostc.state.ak.us) (EVOSTC) was formed following the Exon Valdez oil spill in Alaska in 1989. Since then, then EVOSTC has funded hundreds of projects and in 2012 we began a project to recover and archive the data collected in these EVOSTC funded projects.

For this paper we ask 4 main questions about the data collected from the Exxon Valdez Oil Spill Trustee Council funded projects:

1. Twenty-five years after the EVOS, how many projects funded by EVOSTC can we collect data for?
2. Are there certain **research fields** that are more likely to make data available than others?
3. Are there certain **sectors** that are more likely to make data available than others?
4. Is the availability of data correlated to how old the data are? (temporal relationships?)


#Data setup
*Pull in data and add agency affiliation and minor formatting
*Load required libraries:
  - library(dplyr)
  - library(readr)
  - library(ggpot2)
  - library(reshape)
*Set colors

```{r data prep,echo=FALSE,message=FALSE}

library(dplyr)
library(readr)
library(ggplot2)
library(reshape2)

rslt=read.csv('data/archiveResults2.csv',header=T,stringsAsFactors=F)

rslt$succ=ifelse(rslt$Status %in% c('Published','SentData','Revised'),1,0)

agDes=read.csv('data/ecoInfoAgencyDesig.csv',header=F,stringsAsFactors=T,col.names=c('agency','agGrp'))
agDes$agGrp=gsub('nonProf','private',agDes$agGrp)

rslt2=merge(rslt,agDes,all.x=T)
```
Data table="rslt2"

```{r}
statusCols=c('Unrecoverable'='orangered',
             'Emailed'='orange',
             'Replied'='gold',
             'SentData'='darkolivegreen2',
             'Revised'='darkolivegreen3',
             'Published'='forestgreen')

```

# 1. Project Status Reporting
### Twenty-five years after the EVOS, how many projects funded by EVOSTC can we collect data for?

---
<h4> Plot status of datasets at the end of the initial archiving effort:
```{r status plot,echo=FALSE}

overall=table(rslt2$Status)
overallLong=melt(overall)
colnames(overallLong)=c('Status','nDatasets')
overall2=overallLong %>%
  mutate(prop=nDatasets/sum(nDatasets))

possible=overallLong %>%
  filter(!Status=='Unrecoverable') %>%
  mutate(prop=nDatasets/sum(nDatasets))

statusOrd=c('Unrecoverable','Emailed','Replied','SentData','Revised','Published')
#rslt2$Status=as.factor(rslt2$Status,levels=statusOrd,ordered=T)

statPlot=ggplot()+
  geom_histogram(data=rslt2,aes(x=factor(Status,levels=statusOrd,ordered=T),fill=Status),binwidth=1,order=statusOrd)+
  scale_fill_manual(labels=statusOrd,values=statusCols,breaks=statusOrd)+
  theme(legend.title=element_blank())+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  #theme(axis.text.x = element_text(angle=90))+
  xlab('Status')+
  ylab('Number of datasets\n')+
  ggtitle('Dataset status distribution')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"),
        legend.text = element_text(size = 14, face = "bold"))

statPlot
```


# 2.Field/research categories
### Are there certain **research fields** that are more likely to make data available than others?

# 3. Research center type
### Are there certain **sectors** that are more likely to make data available than others?
(based on PI's affiliation)

# 4. Temporal relationships
### Is the availability of data correlated to how old the data are?