---
title: "EVOS data project - Analysis"
author: "Jessica Couture"
date: "3/15/2017"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: 3
---

## Background

This analysis accompanies the manuscript "A funder imposed data publication requirement seldom inspired data sharing", assessing results of a two year data archiving effort by a group of researchers and students at the National Center for Ecological Analysis and Synthesis at UC Santa Barbara. The [Exxon Valdez Oil Spill Trustee Council](http://www.evostc.state.ak.us) (EVOSTC) was formed following the Exxon Valdez oil spill in Alaska in 1989. Since then, the EVOSTC has funded hundreds of projects and in 2012 an effort was initiated to recover and archive the data collected through these EVOSTC funded projects. The recovery effort spanned two years.

For this paper we use the results of that effort to ask 3 main questions about the data collected from the Exxon Valdez Oil Spill Trustee Council funded projects:

1. Twenty-five years after the EVOS, for how many projects funded by EVOSTC can we collect data?
2. Are there differences in data reporting based on characteristics of the data project?

  * Research field
  * Sector of researching body
  * Year data projects ended

3. Which of these characteristics are most *important* in determining if a dataset will be successfully recovered and how do the important characteristics influence the output (success)?


```{r data prep,echo=FALSE,message=FALSE,warning=F}

library(knitr)
library(dplyr)
library(readr)
library(reshape2)
library(ggplot2)
library(grid)
library(gridExtra)
library(party)

rslt=read.csv('data/evosArchResults.csv',header=T,stringsAsFactors=T)
rslt[rslt$Status=='published','Status']<-'Published'

#rslt$succ=ifelse(rslt$Status %in% c('Published','SentData','Revised'),0,1) # 0=success??? ### RM?
rslt$statSucc=ifelse(rslt$Status %in% c('Published','SentData','Revised'),1,0) # 1 = success, 2 = not successful


agDes=read.csv('data/ecoInfoAgencyDesig.csv',header=T,stringsAsFactors=T,col.names=c('agency','agGrp','agSubGrp'))
agDes$agGrp=gsub('nonProf','private',agDes$agGrp)


rslt2=merge(rslt,agDes,all.x=T)
rslt2=rslt2 %>%
  filter(!end>2010) %>%
  select(Status,start,end,ecosystem,dataType,statSucc,agGrp,agSubGrp,reason)

splBio=strsplit(as.character(rslt2$dataType),'-')

rslt2$dataType<-gsub("biological-",'',rslt2$dataType)

```

## 1. Project Status Reporting
<h4>Twenty-five years after the EVOS, for how many projects funded by EVOSTC can we collect data?</h4>
 
```{r status plot, fig.width=10,echo=FALSE}

overall=table(rslt2$Status)
overallLong=melt(overall)
colnames(overallLong)=c('Status','nDatasets')
overall2=overallLong %>%
  mutate(prop=nDatasets/sum(nDatasets))

possible=overallLong %>%
  filter(!Status=='Unrecoverable') %>%
  mutate(prop=nDatasets/sum(nDatasets))

statusOrd=c('Unrecoverable','Emailed','Replied','SentData','Revised','Published')
stSuccCols=c("0"='gray67',"1"="black")

statPlot=ggplot()+
  geom_bar(data=rslt2,
                 aes(x=factor(Status,levels=statusOrd,ordered=T),fill=factor(statSucc)))+
  scale_x_discrete(labels=c("unrecoverable","outreach\nattempted","contact\nestablished","data\nreceived","published"))+
  scale_fill_manual(name="Data obtained?",
                    breaks=c("1","0"),
                    values=stSuccCols,
                    labels=c("Data recovered", "No data recovered"))+
  theme(legend.title=element_blank())+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  xlab("Recovery stage")+ 
  ylab('Number of projects\n')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"),
        legend.text = element_text(size = 14, face = "bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))

statPlot

```

```{r Effort results}
# Number of projects requested
nProj<-sum(overall)

# percent successful
percRcv<-sum(overall[c("Published","SentData")])/sum(overall)

```
Total number of projects sought = `r nProj`  
Percent success = `r round(percRcv*100)`%

***

## 2. Are there differences in data reporting based on data characteristics?
## Chi-sq tests from Regression

```{r blmChiSq}
blrDat<-rslt2 %>%
  select(end,dataType,statSucc,agSubGrp)

mod<-glm(statSucc~.,family = binomial(link="logit"),data=blrDat)

#rm factors that are not significant

dtMod<-glm(statSucc~dataType,family = binomial(link="logit"),data=blrDat)
#summary(dtMod)

library(aod)

timeChi<-wald.test(b = coef(mod), Sigma = vcov(mod), Terms = 2)

typeChi<-wald.test(b = coef(mod), Sigma = vcov(mod), Terms = 3:11)

agChi<-wald.test(b = coef(mod), Sigma = vcov(mod), Terms = 12:16)

```
Chi-sq for age of data = `r timeChi$result$chi2[[3]][1]`  
Chi-sq for data type = `r typeChi$result$chi2[[3]][1]`  
Chi-sq for agency = `r agChi$result$chi2[[3]][1]`

## Binomial regression

In order to assess how the percent recovery is influenced by time, data type and agency we are running an logistic regression on all 3 factors. 

```{r bionomRegr}

summary(mod)

```


_INTERESING NOTE: When these tests are run on the "start" dates rather than the "end" we get significance in AGE and  DATATYPE_

***
# Vines et al. 2014 style nesting

### Nest 1: Confirmed contact info ("emailed"+)
```{r nestedBLR}
nest<-rslt2 %>%
  mutate(pContInf=ifelse(is.na(reason),1,ifelse(reason=="no contact info",0,1))) %>% #use all date
  mutate(pRepl=ifelse(Status=="Emailed",0,1)) %>% # remove "no contact info" values when analyzing --> of those that we had contact info for, how many replied?
  mutate(pSent=ifelse(Status=="SentData",1,ifelse(Status=="Published",1,0))) %>% # rm "no contact info" & emailed data when analyzing --> of those that 'replied' which sent data?
  mutate(pPub=ifelse(Status=="Published",1,0))

nest1blm<-glm(pContInf~end+dataType+agSubGrp,family = binomial(link="logit"),data=nest)

summary(nest1blm)

summary(nest1b<-glm(pContInf~end,family = binomial(link = "logit"),data=nest))
```
Looking at just how many contact information could be found, there is a significant positive effect of age (p=`r coef(summary(nest1blm))["end",4]`), increasing `r nest1blm$coefficients["end"][[1]]` annually.

### Nest 2: replied given we found contact info ("Replied"+)

```{r nest2}

nest2<-nest %>%
  filter(is.na(reason) | reason !="no contact info")

nest2blm<-glm(pRepl~end+dataType+agSubGrp,family = binomial(link="logit"),data=nest2)

summary(nest2blm)

```

### Nest 3: Sent data given we received a response ("SentData"+)

```{r nest3}

nest3<-nest2 %>%
  filter(!Status=="Emailed")

nest3blm<-glm(pSent~end+dataType+agSubGrp,family = binomial(link="logit"),data=nest3)

summary(nest3blm)

```
Our variables were not significant indicators as to whether data were sent given that we received a response.

### Nest 4: Data were published given we received data ("Published")

```{r nest4}

nest4<-nest3 %>%
  filter(Status %in% c("SentData","Published"))

nest4blm<-glm(pPub~end+dataType+agSubGrp,family = binomial(link="logit"),data=nest4)

summary(nest4blm)

```

Our variables were not significance indicators as to whether data were complete enough to publish given data were sent.

Consistent with Vines, we find that availability of contact information decreases with age of the project. Response, data recovery and ability to publish data are **not** dependent on data or project age. 

***

# 3. Which characteristics are most important in determining if a dataset will be successfully recovered?

We use the "party" package in R to run a random forests analysis to determine which variables are most important. I use the same model as the glm, then create a classification tree below to show _how_ the important variables influence the outcome. This package is be better than the "randomForest" package when independent variables are different types (Strobl et al. 2009).

For the random forests the independent variable with the highest absolute value has the highest impact on the dependent variable.

### Random forests
```{r randForestBio, message=F}
rslt2$dataType<-as.factor(rslt2$dataType)
partyForBio<-cforest(statSucc~agSubGrp+end+dataType,data=rslt2,controls = cforest_unbiased(mtry = 2, ntree = 1000))
varimp(partyForBio)
```

__Based on these results the most important variable in determining the outcome is research field__

***

## How do the important characteristics influence the output (success)? 
### Classification tree

Here we run one iteration of the forest analysis above to display which variables whithin each classification determine positive or negative results.

```{r class tree bio, message=F, warning=FALSE}
partreeBio<-ctree(statSucc~agSubGrp+end+factor(dataType),data=rslt2)

partreeBio

plot(partreeBio)
```

_Birds, fish, habitat, mammal_ and _modeling data_ result in negative results.
_Benthic invertebrates, plankton, oil,_ and _physical data_ result in positive results.