---
title: "Regression analysis"
author: "Jessica Couture"
date: "March 9, 2018"
output: html_document
---
## Background

This analysis accompanies the manuscript "Funder imposed data publication requirements seldom inspire data sharing", assessing results of a two year data archiving effort by a group of researchers and students at the National Center for Ecological Analysis and Synthesis at UC Santa Barbara. The [Exxon Valdez Oil Spill Trustee Council](http://www.evostc.state.ak.us) (EVOSTC) was formed following the Exxon Valdez oil spill in Alaska in 1989. Since then, the EVOSTC has funded hundreds of projects and in 2012 an effort was initiated to recover and archive the data collected through these EVOSTC funded projects. The recovery effort spanned two years.

For this paper we ask 3 main questions about the data collected from the Exxon Valdez Oil Spill Trustee Council funded projects:

1. Twenty-five years after the EVOS, for how many projects funded by EVOSTC can we collect data?
2. Are there differences in data reporting based on characteristics of the data?

  * Research field
  * Sector of researching body
  * Year data projects ended

3. Which of these characteristics are most *important* in determining if a dataset will be successfully recovered and how do the important characteristics influence the output (success)?

```{r data prep,echo=FALSE,message=FALSE,warning=F}

library(knitr)
library(dplyr)
library(readr)
library(ggplot2)
library(reshape2)
library(grid)
library(gridExtra)
library(party)

rslt=read.csv('data/evosArchResults.csv',header=T,stringsAsFactors=T)
rslt[rslt$Status=='published','Status']<-'Published'

rslt$succ=ifelse(rslt$Status %in% c('Published','SentData','Revised'),0,1) # 0=success??? ### RM?
rslt$statSucc=ifelse(rslt$Status %in% c('Published','SentData','Revised'),1,0) # 1 = success, 2 = not successful


agDes=read.csv('data/ecoInfoAgencyDesig.csv',header=T,stringsAsFactors=T,col.names=c('agency','agGrp','agSubGrp'))
agDes$agGrp=gsub('nonProf','private',agDes$agGrp)


rslt2=merge(rslt,agDes,all.x=T)
rslt2=rslt2 %>%
  filter(!end>2010) %>%
  select(Status,start,end,ecosystem,dataType,statSucc,agGrp,agSubGrp,reason)

splBio=strsplit(as.character(rslt2$dataType),'-')

rslt2$dataType<-gsub("biological-",'',rslt2$dataType)

```

## Binomial regression

In order to assess how the percent recovery is influenced by time, data type and agency we are running an logistic regression on all 3 factors. 

```{r binomRegr}
blrDat<-rslt2 %>%
  select(end,dataType,statSucc,agSubGrp)

mod<-glm(statSucc~.,family = binomial(link="logit"),data=blrDat)

summary(mod)

ageMod<-glm(statSucc~end-1,family = binomial(link="logit"),data=blrDat)
#summary(ageMod)
```

# Chi-sq tests from Regression:

```{r chiSq}
library(aod)

timeChi<-wald.test(b = coef(mod), Sigma = vcov(mod), Terms = 2)

typeChi<-wald.test(b = coef(mod), Sigma = vcov(mod), Terms = 3:11)

agChi<-wald.test(b = coef(mod), Sigma = vcov(mod), Terms = 12:16)

```

Chi-sq for age of data = `r timeChi$result$chi2[[3]]`

Chi-sq for data type = `r typeChi$result$chi2[[3]]`

Chi-sq for agency = `r agChi$result$chi2[[3]]`

These chi-sq results match what we found before: only data-type is significant
___

## Binomial regression: "START"

Playing with the data I saw that then stats are run using "start" instead of "end" for age of data, we get a lot more significance. I know we decided on "end" for a reason but might be worth discussing in our group and maybe adding to the discussion. 

```{r binomRegr-Start}
strtDat<-rslt2 %>%
  select(start,dataType,statSucc,agSubGrp)

modSt<-glm(statSucc~.,family = binomial(link="logit"),data=strtDat)

summary(modSt)

ageMod<-glm(statSucc~start-1,family = binomial(link="logit"),data=strtDat)
#summary(ageMod)
```

# Chi-sq tests from Regression:

```{r chiSq-start}
library(aod)

timeChiSt<-wald.test(b = coef(modSt), Sigma = vcov(modSt), Terms = 2)

typeChiSt<-wald.test(b = coef(modSt), Sigma = vcov(modSt), Terms = 3:11)

agChiSt<-wald.test(b = coef(modSt), Sigma = vcov(modSt), Terms = 12:16)

```

Chi-sq for age of data = `r timeChiSt$result$chi2[[3]]`

Chi-sq for data type = `r typeChiSt$result$chi2[[3]]`

Chi-sq for agency = `r agChiSt$result$chi2[[3]]`

_Using "start" the age of data is in fact SIGNIFICANT along with data type_

```{r chi squared - end years,echo=FALSE,warning=FALSE}

startProps=table(rslt2$statSucc,as.character(rslt2$start))
startProps

strtChi=chisq.test(startProps)

strtChi

```

## to dos:

mimic Vines' status analyese (see paper, last paragraph) --> don't think this is relevant for our analysis

include other categories individually--> ditto

