---
title: "How do we want to display TEMPORAL data?"
author: "Jessica Couture"
date: "5/22/2016"
output: html_document
---
```{r setup, include=FALSE}
library(knitr)
library(plyr)
library(dplyr)
library(readr)
library(ggplot2)
library(reshape2)
library(grid)
library(gridExtra)
library(gtable)

rslt=read.csv('data/archResRevised2.csv',header=T,stringsAsFactors=F)
rslt[rslt$Status=='published','Status']<-'Published'

rslt$succ=ifelse(rslt$Status %in% c('Published','SentData','Revised'),'aff','neg')
rslt$plotSucc=ifelse(rslt$Status %in% c('Published','SentData','Revised'),1,2)# 1 = success, 2 = not successful

rslt$statSucc=ifelse(rslt$Status %in% c('Published','SentData','Revised'),1,0)# 1 = success, 0 = not successful


agDes=read.csv('data/ecoInfoAgencyDesig.csv',header=T,stringsAsFactors=T,col.names=c('agency','agGrp','agSubGrp'))
agDes$agGrp=gsub('nonProf','private',agDes$agGrp)

rslt2=merge(rslt,agDes,all.x=T)
```

## Emphasize percents? number of datasets? differences between years?...
We have talked about a number of different ways to display these results. Here I show 2 of the variables discussed:

* Look at all years projects were funded or just the last year?
* Use percent successful or number of datasets?

Below I have plotted first the data looking at only the last year of funding (so these would represent the amount of time since a project was finished) as raw number of datasets and percents, then look at the data collected for each year a dataset was funded. I can think of pros and cons for each so this needs to be discussed

I didn't start with stats because I haven't run any, something else we need to discuss....

> Q4. Is the availability of data correlated to how old the data are? (temporal relationships?)

#### __Plots of data reporting based on the year project funding _ended,_ to represent length of time since data were created:__
```{r temporal-endYr, echo=FALSE}

rslt2$succ<-ifelse(rslt2$Status %in% c('Unrecoverable','Emailed','Replied'),1,0)

endYr=rslt2 %>%
  filter(!end<1989) %>%
  filter(!end>2010) %>%
  select(succ,end)

endYrs=as.data.frame(table(endYr$end,endYr$succ))
colnames(endYrs)=c('year','succ','freq')
endYrs<- mutate(endYrs,since=(2012-as.numeric(paste(endYrs$year))))

yrsCols=c("1"='grey',
          "0"='black')

ggplot(endYrs,aes(x=since,y=freq,fill=succ,order=rev(succ)))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values=yrsCols,name='Success?',breaks=c('1','0'),labels=c('no','yes'))+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  xlim(max(endYrs$since),min(endYrs$since))+
  xlab('Years since project ended')+
  ylab('Number of datasets\n')+
  ggtitle('Number of EVOS datasets collected\nbased on the number of years since a project ended')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))
```

***

``` {r temporal percents - end, echo=F}

yrsPropEnd<-endYrs %>%
  group_by(as.character(year)) %>%
  mutate(yrSum=sum(freq)) %>%
  mutate(prop=freq/yrSum)

yrsPercEnd=ggplot(yrsPropEnd[yrsPropEnd$succ==0,],aes(x=since,y=prop,))+
  geom_bar(stat = "identity")+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  xlim(max(endYrs$since),min(endYrs$since))+
  ylim(0,1)+
  xlab('Year')+
  ylab('Percent successful')+
  ggtitle('Percent success based on the\nlast year a project was funded')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))
yrsPercEnd
```

***

#### __These plots show results based on each year a project received funding__
Ex: a project funded from 1994-1997 would be counted 4 times: 1994, 1995, 1996, 1997
```{r temporal, echo=F}
library(plyr)
yrsList=list()
placer=1
for(i in 1:nrow(rslt2)) {
  yrsList[[i]]=as.numeric(rslt2$start[i]):as.numeric(rslt2$end[i])
  placer=placer+1
}

yrsDf=ldply(yrsList,rbind)
yrsDf$status=rslt2$Status
yrsDf$category=rslt2$category
colnames(yrsDf)=c(paste('yrs',colnames(yrsDf)[1:89],sep='.'),'status','category')

yrsLng=reshape(yrsDf,varying=c(1:89),direction='long')
yrs3=yrsLng %>%
  filter(!is.na(yrs)) %>% 
  filter(!yrs<1989) %>%
  filter(!yrs>2010) %>%
  mutate(succ=ifelse(status %in% c('Published','SentData','Revised'),0,1)) %>%
  select(succ,category,yrs)

totYrs=as.data.frame(table(yrs3$yrs,yrs3$succ))
colnames(totYrs)=c('year','succ','freq')

totYrsAll<-ggplot(totYrs,aes(x=year,y=freq,fill=succ,order=rev(succ)))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values=yrsCols,name='Success?',breaks=c('neg','aff'),labels=c('no','yes'))+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  theme(axis.text.x = element_text(angle=90))+
  xlab('Year')+
  ylab('Number of datasets\n')+
  ggtitle('Number of EVOS datasets collected\nbased on year data were funded')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black")) ### GOOD!
totYrsAll
```

***

``` {r temporal percents - each year, echo=F,warnings=F}
detach(package:plyr, unload = TRUE)
library(dplyr)
yrsProp<-totYrs %>%
  group_by(as.character(year)) %>%
  mutate(yrSum=sum(freq)) %>%
  mutate(prop=freq/yrSum)

yrsPercEach=ggplot(yrsProp[yrsProp$succ==0,],aes(x=year,y=prop,))+
  geom_bar(stat = "identity")+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  theme(axis.text.x = element_text(angle=90))+
  ylim(0,1)+
  xlab('Year')+
  ylab('Percent successful')+
  ggtitle('Percent success by year')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))
yrsPercEach
```

***

#### __Combining the plots together:__

```{r overlay yrs plot - end, echo=F}

p2=ggplot(yrsPropEnd[yrsPropEnd$succ==0,],aes(x=since,y=prop))+
  geom_line(stat = "identity",colour='red',size=1)+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill=NA),
        panel.background = element_blank())+
  xlim(max(endYrs$since),min(endYrs$since))+
  ylim(0,1)+
  #xlab('Years since funding award')+
  ylab('Success rate')+
  #ggtitle('Percent over time')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))


p1 <- ggplot(yrsPropEnd,aes(x=since,y=freq,fill=succ,order=rev(succ)))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values=yrsCols,name='Success?',breaks=c('1','0'),labels=c('no','yes'),guide=F)+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  xlim(max(endYrs$since),min(endYrs$since))+
  xlab('Years since project ended')+
  ylab('Number of datasets\n')+
  #ggtitle('Number of EVOS datasets collected\nbased on the number of years since a project ended')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))

# extract gtable
g1 <- ggplot_gtable(ggplot_build(p1))
g2 <- ggplot_gtable(ggplot_build(p2))

# overlap the panel of 2nd plot on that of 1st plot
pp <- c(subset(g1$layout, name == "panel", se = t:r))
g <- gtable_add_grob(g1, g2$grobs[[which(g2$layout$name == "panel")]], pp$t, 
                     pp$l, pp$b, pp$l)

# axis tweaks
ia <- which(g2$layout$name == "axis-l")
ga <- g2$grobs[[ia]]
ax <- ga$children[[2]]
ax$widths <- rev(ax$widths)
ax$grobs <- rev(ax$grobs)
ax$grobs[[1]]$x <- ax$grobs[[1]]$x - unit(1, "npc") + unit(0.15, "cm")
g <- gtable_add_cols(g, g2$widths[g2$layout[ia, ]$l], length(g$widths) - 1)
g <- gtable_add_grob(g, ax, pp$t, length(g$widths) - 1, pp$b)

# draw it
grid.draw(g)
```

***

#### __Two plots:__
```{r stacked yrs plots - end, fig.width=6, fig.height=7,echo=F}
stk2=ggplot(yrsPropEnd[yrsPropEnd$succ==0,],aes(x=since,y=round(prop*100)))+
  geom_line(stat = "identity",colour='red',size=1)+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  xlim(max(endYrs$since),min(endYrs$since))+
  ylim(0,100)+
  ylab('Success rate (%)\n')+
  theme(axis.text.y=element_text(size=14),
        title=element_text(size=16,face="bold"),
        axis.title.x = element_blank(),
        axis.text.x  = element_blank())+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))


stk1 <- ggplot(yrsPropEnd,aes(x=since,y=freq,fill=succ,order=rev(succ)))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values=yrsCols,name='Success?',breaks=c('1','0'),labels=c('no','yes'),guide=F)+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  xlim(max(endYrs$since),min(endYrs$since))+
  xlab('Years since project ended')+
  ylab('Number of datasets\n')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))

grid.arrange(stk2, stk1, ncol = 1, heights = c(1,2))

```


```{r stacked yrs plots - each year, fig.width=6, fig.height=7, echo=F}
stkEa2=ggplot(yrsProp[yrsProp$succ==0,],aes(x=as.numeric(as.character(year)),y=(prop*100)))+
    geom_line(stat = "identity",colour='red',size=1)+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  ylim(0,100)+
  ylab('Success rate (%)\n')+
  theme(axis.text.y=element_text(size=14),
        title=element_text(size=16,face="bold"),
        axis.title.x = element_blank(),
        axis.text.x  = element_blank())+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))


stkEa1 <- ggplot(yrsProp,aes(x=year,y=freq,fill=succ,order=rev(succ)))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values=yrsCols,name='Success?',breaks=c('1','0'),labels=c('no','yes'),guide=F)+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank())+
  xlab('Year of project funding')+
  ylab('Number of datasets\n')+
  #ggtitle('Number of EVOS datasets collected\nbased on the number of years since a project ended')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"),
        axis.text.x  = element_text(angle=90, vjust=0.5))+
  theme(axis.line.x = element_line(color="black"),
        axis.line.y = element_line(color="black"))

grid.arrange(stkEa2, stkEa1, ncol = 1, heights = c(1,2))
```


### side-by-side bars

```{r sideBySideBars, echo=F}

grpd<-yrsPropEnd
grpd$prop[23:44]<-0
grpd$yrSum[23:44]<-0

propBar=ggplot(grpd,aes(x=since,y=(prop*100),fill = succ)) +
  geom_bar(position = "dodge", stat="identity")+
  scale_fill_manual(values = c("red", "red"))+
  #scale_x_continuous(breaks = 1:10)+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black",fill=NA),
        panel.background = element_blank(),
        axis.title.y=element_text(color="red"))+
  xlim(max(endYrs$since),min(endYrs$since))+
  ylim(0,100)+ 
  #xlab('Years since funding award')+
  ylab('Success rate')+
  #ggtitle('Percent over time')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"),
        legend.position="none")


nBar <- ggplot(grpd,aes(x=since,y=yrSum,fill = rev(succ)))+
  geom_bar(position = "dodge", stat="identity")+
  scale_fill_manual(values = c("blue", "blue"))+
  theme(axis.line=element_line('black'),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.title.y=element_text(color="blue"))+
  xlim(max(endYrs$since),min(endYrs$since))+
  xlab('Years since project ended')+
  ylab('Number of datasets')+
  #ggtitle('Number of EVOS datasets collected\nbased on the number of years since a project ended')+
  theme(axis.text=element_text(size=14),
        title=element_text(size=16,face="bold"),
        legend.position="none")

# extract gtable
pb <- ggplot_gtable(ggplot_build(nBar))
nb <- ggplot_gtable(ggplot_build(propBar))

# overlap the panel of 2nd plot on that of 1st plot
pp <- c(subset(pb$layout, name == "panel", se = t:r))
g <- gtable_add_grob(pb, nb$grobs[[which(nb$layout$name == "panel")]], pp$t, 
                     pp$l, pp$b, pp$l)

# axis tweaks
ia <- which(nb$layout$name == "axis-l")
ga <- nb$grobs[[ia]]
ax <- ga$children[[2]]
ax$widths <- rev(ax$widths)
ax$grobs <- rev(ax$grobs)
ax$grobs[[1]]$x <- ax$grobs[[1]]$x - unit(1, "npc") + unit(0.15, "cm")
g <- gtable_add_cols(g, nb$widths[nb$layout[ia, ]$l], length(g$widths) - 1)
g <- gtable_add_grob(g, ax, pp$t, length(g$widths) - 1, pp$b)
g <- gtable_add_grob(g, g2$grob[[7]], pp$t, length(g$widths), pp$b)
# draw it
grid.draw(g)
```