---
title: "EcoInfo paper category re-organization"
author: "Jessica Couture"
date: "4/23/2016"
output: html_document
---
```{r echo=FALSE,message=FALSE}
library(dplyr)
```

#### This script re-organizes the 400+ datasets we have for the EVOSTC project into more sensible groups showing a bit of the process for feedback:

***

## **Original mess:**

##### We started with the following category structure which intially was based on a system and got more and more random as I came across projects that didn't fit within my original categories:

* modeling
* habitat
* fish
* dataMgmt
* monitoring
* terrestrial
* inverts
* oceanography
* marineBirds
* mammals
* anthropogenic
* oil
* plankton
* mapping
* freshWater
* meta
* misc projects
* unk

```{r, echo=FALSE,fig.height=5}
dat=read.csv('archResNewCateg.csv',header=T,stringsAsFactors =F,na.strings=c('',' '))

datFill<-dat[is.na(dat$category.1),1:13]

par(mar = c(11,4,4,2) + 0.1)
barplot(table(datFill$category),
        las=3)
```

...Kinda all over the place.

***

## **First re-organization:**

##### When Rachael, Colette and I got together to make new groups, we came up with the following:

| DataType        | Sub-category  |
| --------------- | ------------- |
| Biological      | fish          |
| Biological      | benthicInverts|
| Biological      | mammals       |
| Biological      | plankton      |
| Physical        | NA            |
| Chemical        | NA            |
| Other (default) | NA            |

##### This still leaves us with a lot of "other" since everything doesn't fit nicely into these clean groups:

```{r, results='hide',echo=F,message=FALSE}
require(dplyr)
datFill <- datFill %>%
  mutate(dataType = ifelse(datFill$category=='oceanography','physical',
                           ifelse(datFill$category=='oil','chemical',
                                  ifelse(datFill$category=='fish','biological-fish',
                                         ifelse(datFill$category=='mammals','biological-mammals',
                                                ifelse(datFill$category=='marineBirds','biological-birds',
                                                       ifelse(datFill$category=='inverts','biological-benthicInverts',
                                                              ifelse(datFill$category=='plankton','biological-plankton','other'))))))))
         
datFill[93,'dataType']<-'ecosystem/habitat' #Eric Meyers
datFill[93,'ecosystem']<-'terrestrial/freshWater'
datFill[120,'dataType']<-'ecosystem/habitat' #Geoffrey Coble
datFill[120,'ecosystem']<-'terrestrial/freshWater'
datFill[138,'dataType']<-'biological-mammals' #james Faro
datFill[138,'ecosystem']<-'terrestrial/freshWater'
datFill[148,'dataType']<-'biological-birds' #Jeffrey Hughes
datFill[148,'ecosystem']<-'terrestrial/freshWater'
datFill[167,'dataType']<-'biological-fish' #Jim Edmondson
datFill[167,'ecosystem']<-'terrestrial/freshWater'
datFill[186,'dataType']<-'biological-mammals' #Jonathan Lewis
datFill[186,'ecosystem']<-'terrestrial/freshWater'
datFill[201,'dataType']<-'ecosystem/habitat' #Ken Hodges
datFill[201,'ecosystem']<-'terrestrial/freshWater'
datFill[253,'dataType']<-'biological-mammals' #Merav Ben-David
datFill[253,'ecosystem']<-'terrestrial/freshWater' 
datFill[273,'dataType']<-'biological-mammals' #PI unavailable, black bear study
datFill[273,'ecosystem']<-'terrestrial/freshWater' 
datFill[274,'dataType']<-'ecosystem/habitat' #PI unavailable
datFill[274,'ecosystem']<-'terrestrial/freshWater'
datFill[297,'dataType']<-'biological-mammals' #Richard Sellers
datFill[297,'ecosystem']<-'terrestrial/freshWater' 
datFill[298,'dataType']<-'biological-mammals' #Richard Sellers
datFill[298,'ecosystem']<-'terrestrial/freshWater'
datFill[302,'dataType']<-'biological-mammals' #Robert White
datFill[302,'ecosystem']<-'terrestrial/freshWater' 
datFill[330,'dataType']<-'biological-mammals' #Steven Jewette
datFill[330,'ecosystem']<-'terrestrial/freshWater' 
datFill[340,'dataType']<-'physical' #Sue Mauger
datFill[340,'ecosystem']<-'terrestrial/freshWater' 
datFill[351,'dataType']<-'biological-birds' #Tim Bowman
datFill[351,'ecosystem']<-'terrestrial/freshWater' 
datFill[352,'dataType']<-'biological-birds' #Tim Bowman
datFill[352,'ecosystem']<-'terrestrial/freshWater' 
datFill[364,'dataType']<-'biological-birds' #Vernon Byrd
datFill[364,'ecosystem']<-'terrestrial/freshWater'
```


```{r, echo=FALSE,message=FALSE}
par(mar = c(11,4,4,2) + 0.1)
barplot(table(factor(datFill$dataType,levels=c('physical',
                                               'chemical',
                                               'biological-plankton',
                                               'biological-benthicInverts',
                                               'biological-fish',
                                               'biological-birds',
                                               'biological-mammals',
                                               'other'
                                               ))),
        las=3)
        text(9.15,sum(datFill$dataType=='other')+7,as.character(sum(datFill$dataType=='other')))
```

***

## **Adding an "ecosystem/habitat" group:**

##### Rachael and I briefly discussed the many modeling, mapping and habitat projects that were lumped into the "other" category we decided that pulling these out into an "ecosystem/habitat" category might be a little vague but would work for these groups. Here's what this does to the distribution of datasets:

```{r, echo=FALSE}
datFill$dataType=ifelse(datFill$category %in% c('habitat','modeling','mapping'), 'ecosystem/habitat',datFill$dataType)
par(mar = c(11,4,4,2) + 0.1)
barplot(table(factor(datFill$dataType,levels=c('physical',
                                               'chemical',
                                               'biological-plankton',
                                               'biological-benthicInverts',
                                               'biological-fish',
                                               'biological-birds',
                                               'biological-mammals',
                                               'ecosystem/habitat',
                                               'other'
                                               ))),
        las=3)
        text(10.15,sum(datFill$dataType=='other')+7,as.character(sum(datFill$dataType=='other')))
```

##### This brings the other group down to 22 datastes, which can be loosely described by the following terms:
(each row is a dataset with 1-2 descriptive terms)

```{r, results='asis',echo=FALSE}
        
subCat<-c('projMgmt','projMgmt','economics','projMgmt','economics','subsistence','communityMgmt','social','environmentalImpacts','samplingImpacts','projMgmt','meeting','economics','economics','economics','economics','economics','economics','subsistence','dataMgmt','dataMgmt','communityMgmt')
subCat2<-c('habitat','ecosystem','recreation','habitat','subsistence', 'film','waste','subsistence','infrastructure','metaAnalysis',NA,'10yrSymposium','resarchPrograms','subsistence','fisheries','resources','resources','fisheries','resources',NA,NA,'waste')
descriptors=data.frame(subCat=subCat,subCat2=subCat2)

tableCat <- function(inFrame) {
    outText <- paste(names(inFrame), collapse = " | ")
    outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse = " | "))
    invisible(apply(inFrame, 1, function(inRow) {
        outText <<- c(outText, paste(inRow, collapse = " | "))
    }))
    return(outText)
}
otherTabInsert<-tableCat(descriptors)
cat(otherTabInsert, sep = "\n")
```

###### **Note:** I haven't sorted these into "terrestrial", "nearshore" and "pelagic" yet. That will take a bit more work.


```{r removing no-EVOSTC}
arr=read.csv("data/archResRevised2.csv",header = T)
arr2=arr %>%
  filter(!dataEditNotes=="nonEvos?") %>%
  mutate(project=paste(Status,parentTask,sep="-")) %>%
  filter(!duplicated(project)) # 22 projects have data in multiple status'

dupes=arr2[duplicated(arr2$parentTask),"parentTask"]
dupeDf=arr2[arr2$parentTask %in% dupes,]
dupPub<-dupeDf[dupeDf$Status %in% c("Published","SentData"),]
dupNos<-dupeDf%>%
  filter(!dupeDf$parentTask %in% dupPub$parentTask)%>%
  filter(!duplicated(parentTask))
dupNos$Status=c("Unrecoverable","Replied","Replied")

arr3=arr2 %>%
  filter(!parentTask %in% dupes)
arr4=rbind(arr3,dupPub,dupNos)

## reorganize dataType categories

levels(arr4$dataType)=c(levels(arr4$dataType),"oil","biological-habitat","modeling")
arr4$dataType[arr4$category=="oil"]<-"oil"
arr4$dataType[arr4$category=="modeling"]<-"modeling"
arr4$dataType[arr4$category=="habitat"]<-"biological-habitat"
arr4$dataType[arr4$category=="terrestrial" & arr4$dataType=='ecosystem/habitat']<-"biological-habitat"
arr4$dataType[arr4$category=="mapping" & arr4$dataType=='ecosystem/habitat']<-"biological-habitat"
arr4$dataType<-gsub("chemical","physical",arr$dataType)

#write.csv(arr4,file="data/archResRevised2.csv",row.names=F)

```